Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs

### 2021-02-22 submission/manuscript.Rmd
Fourth round of revisions

single_copy <- rrn_copy_number %>% filter(median_rrn == 1) %>% top_n(n=1, N)
max_copy <- rrn_copy_number %>% top_n(n=1, median_rrn)

@@

16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985]. Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds date to at least the 1990s. The distance-based thresholds that were developed and are now widely used (i.e., 3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007]. Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018]. As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014]. Approaches for identifying ASVs do not cluster sequences based on a distance-based threshold [@Callahan2017]. Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than one copy of the *rrn* operon and those copies are not identical [@Pei2010; @Sun2013]. Yet, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple clusters. Conversely, using too broad of a threshold to define OTUs creates the risk of clustering together bacterial species into the same OTU. An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S. epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene in these two genomes is distinct and represent 10 ASVs. Conversely, if the copies were clustered using a 3% distance threshold, then all 10 ASVs would cluster into the same OTU. The goal of this study was to quantify the risk of splitting a single genome into multiple clusters and the risk of clustering different bacterial species into the same cluster when using ASVs and various OTU definitions.

To investigate the variation in the number of copies of the 16S rRNA gene per genome and the intragenomic variation among copies of the 16S rRNA gene, I obtained 16S rRNA sequences from the *rrn* copy number database (*rrn*DB)[@Stoddard2014]. Among the `r n_distinct(metadata$species)` species represented in the *rrn*DB there were `r nrow(metadata)` genomes. The median *rrn* copy number per species ranged between `r single_copy$median_rrn` (e.g., *`r single_copy$species`*) and `r max_copy$median_rrn` (*`r max_copy$species`*). As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased. On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v34`, `r rate_v4`, and `r rate_v45` variants when considering the V3-V4, V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**). For example, the `r single_copy_N` genome accessions of *`r single_copy$species`* in the *rrn*DB each had `r single_copy$median_rrn` copy of the gene per genome. However, across those accessions, there were `r single_nasvs` versions of the gene. An *E. coli* genome typically had `r ecoli_copy$median_rrn` copies of the 16S rRNA gene with a median of `r ecoli_copy$median_asv` distinct full length ASVs per genome (intraquartile range between `r ecoli_copy$l_quartile_asv` and `r ecoli_copy$u_quartile_asv`). Across the `r ecoli_copy$N` *E. coli* genomes in the *rrn*DB, there were `r ecoli_nasvs` versions of the gene. These observations highlight the risk of selecting a threshold for defining clusters that is too narrow because it is possible to split a single genome into multiple clusters.

A method to avoid splitting a single genome into multiple clusters is to cluster 16S rRNA gene sequences together based on their distances between each other. Therefore, I assessed the impact of the distance threshold used to define clusters of 16S rRNA genes on the propensity to split a genome into separate clusters. To control for uneven representation of genomes across species, I randomly selected one genome from each species for the following analyses and repeated each randomization 100 times. I observed that as the *rrn* copy number increased, the distance threshold required to reduce the ASVs in each genome to a single OTU increased (Figure 1). Among species with 7 copies of the *rrn* operon (e.g., *E. coli*), a distance threshold of `r format_pct(threshold_v19)`% was required to reduce full length ASVs into a single OTU for 95% of the species. Similarly, thresholds of `r format_pct(threshold_v34)`, `r format_pct(threshold_v4)`, and `r format_pct(threshold_v45)`% were required for the V3-V4, V4, and V4-V5 regions, respectively. But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively. Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters. At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU (**Figure 2**). Using ASVs, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. Considering that species designations are unevenly applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs more problematic than clustering species together. Therefore, larger thresholds are advisable.

@@

**(vi) Note on usage of ASV, OTU, and cluster.** I used "ASV" to denote the cluster of true 16S rRNA gene sequences that were identical to each other and "OTU" to denote the product of distance-based clustering of sequences. Although ASVs do represent a type of operational defition of a taxonomic unit and can be thought of as an OTU formed using a distance of zero, proponents of the ASV approach prefer to avoid the term OTU given the long history of OTUs being formed by distance-based clustering (https://github.com/benjjneb/dada2/issues/62; accessed 2021-02-19). For this reason, when an ASV splits a genome into different units those units are called clusters rather than OTUs.

@@

**Acknowledgements.** I am grateful to Robert Hein and Thomas Schmidt, who maintain the *rrn*DB, for their help in understanding the curation of the database and for making the 16S rRNA gene sequences and related metadata publicly available. I am also grateful to community members who watched the serialized version of this analysis on YouTube and provided suggestions and questions over the course of the development of this project. This work was supported, in part, through grants from the NIH to PDS (P30DK034933, U01AI124255, and R01CA215574).

@@

**Figure S1. The ratio of number of distinct ASVs per copy of the *rrn* operon increased for a species as the number of genomes in the *rrn*DB for that species increased.** Each point represents a different species and was shaded to be 80% transparent so that when points overlap they become darker. The blue line represents a smoothed fit through the data. The axes both use a logarithmic scale (base 10).

>>>>

single_copy <- rrn_copy_number %>% filter(median_rrn == 1) %>% top_n(n=1, N)

stopifnot(single_copy$min_rrn == single_copy$max_rrn)

max_copy <- rrn_copy_number %>% top_n(n=1, median_rrn)

@@

16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985]. Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds date to at least the 1990s. The distance-based threshold that was developed and is now widely used was based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007]. Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018]. As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014]. Approaches for identifying ASVs do not cluster sequences based on a distance-based threshold [@Callahan2017]. Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than one copy of the *rrn* operon and that those copies are not identical [@Pei2010; @Sun2013]. Yet, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple clusters. Conversely, using too broad of a threshold to define OTUs creates the risk of clustering together bacterial species into the same OTU. An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S. epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene in these two genomes is distinct and represent 10 ASVs. Conversely, if the copies were clustered using a 3% distance threshold, then all 10 ASVs would cluster into the same OTU. The goal of this study was to quantify the tradeoff of splitting a single genome into multiple clusters and the risk of clustering different bacterial species into the same cluster when using ASVs and various OTU definitions.

To investigate the variation in the number of copies of the 16S rRNA gene per genome and the intragenomic variation among copies of the 16S rRNA gene, I obtained 16S rRNA sequences from the *rrn* copy number database (*rrn*DB)[@Stoddard2014]. Among the `r n_distinct(metadata$species)` species represented in the *rrn*DB there were `r nrow(metadata)` genomes. The median *rrn* copy number per species ranged between `r single_copy$median_rrn` (e.g., *`r single_copy$species`*) and `r max_copy$median_rrn` (*`r max_copy$species`*). As **rrn** copy number for a genome increased, the number of variants of the 16S rRNA gene in each genome also increased. On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v34`, `r rate_v4`, and `r rate_v45` variants when considering the V3-V4, V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**). For example, the `r single_copy_N` genome accessions of *`r single_copy$species`* in the *rrn*DB each had `r single_copy$median_rrn` copy of the gene per genome. However, across those accessions, there were `r single_nasvs` versions of the gene. An *E. coli* genome typically had `r ecoli_copy$median_rrn` copies of the 16S rRNA gene with a median of `r ecoli_copy$median_asv` distinct full length ASVs per genome (intraquartile range between `r ecoli_copy$l_quartile_asv` and `r ecoli_copy$u_quartile_asv`). Across the `r ecoli_copy$N` *E. coli* genomes in the *rrn*DB, there were `r ecoli_nasvs` versions of the gene. These observations highlight the risk of selecting a threshold for defining clusters that is too narrow because it is possible to split a single genome into multiple clusters.

A method to avoid splitting a single genome into multiple clusters is to cluster 16S rRNA gene sequences together based on their distances between each other. Therefore, I assessed the impact of the distance threshold used to define clusters of 16S rRNA genes on the propensity to split a genome into separate clusters. To control for uneven representation of genomes across species, I randomly selected one genome from each species for the following analyses and repeated each randomization 100 times. I observed that as the *rrn* copy number increased, the distance threshold required to reduce the ASVs in each genome to a single OTU increased (Figure 1). Among species with 7 copies of the *rrn* operon (e.g., *E. coli*), a distance threshold of `r format_pct(threshold_v19)`% was required to reduce full length ASVs into a single OTU for 95% of the species (Figure 2). Similarly, thresholds of `r format_pct(threshold_v34)`, `r format_pct(threshold_v4)`, and `r format_pct(threshold_v45)`% were required for the V3-V4, V4, and V4-V5 regions, respectively. But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 2). Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters. At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU (**Figure 2**). Using ASVs, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. Considering that species designations are unevenly applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs is more problematic than clustering species together. Therefore, larger thresholds are advisable.

@@

**(vi) Note on usage of ASV, OTU, and cluster.** I used "ASV" to denote the cluster of true 16S rRNA gene sequences that were identical to each other and "OTU" to denote the product of distance-based clustering of sequences. Although ASVs do represent a type of operational defition of a taxonomic unit and can be thought of as an OTU formed using a distance of zero, proponents of the ASV approach prefer to avoid the term OTU given the long history of OTUs being formed by distance-based clustering (https://github.com/benjjneb/dada2/issues/62; accessed 2021-02-19). For this reason, when an ASV splits a genome into different units, those units were called clusters rather than OTUs.

@@

**Acknowledgements.** I am grateful to Robert Hein and Thomas Schmidt, who maintain the *rrn*DB, for their help in understanding the curation of the database and for making the 16S rRNA gene sequences and related metadata publicly available. I am also grateful to community members who watched the serialized version of this analysis on YouTube and provided suggestions and questions over the course of the development of this project. This work was supported, in part, through grants from the NIH (P30DK034933, U01AI124255, and R01CA215574).

@@

**Figure S1. The ratio of number of distinct ASVs per copy of the *rrn* operon increased for a species as the number of genomes in the *rrn*DB for that species increased.** Each point represents a different species and was shaded to be 80% transparent so that when points overlap they become darker. The blue line represents a smoothed fit through the data. Both axes use a logarithmic scale (base 10).

### 2021-02-22 submission/manuscript.tex
Fourth round of revisions

16S rRNA gene sequencing is a powerful technique for describing and
comparing microbial communities (1). Efforts to link 16S rRNA gene
sequences to taxonomic levels based on distance thresholds date to at
least the 1990s. The distance-based thresholds that were developed and
are now widely used (i.e., 3\%) were based on DNA-DNA hybridization
approaches that are not as precise as genome sequencing (2, 3). Instead,
genome sequencing technologies have suggested that the widely used 3\%
distance threshold to operationally define bacterial taxa is too coarse
(4--6). As an alternative to OTUs, amplicon sequencing variants (ASVs)
have been proposed as a way to adopt the thresholds suggested by genome
sequencing to microbial community analysis using 16S rRNA gene sequences
(7--10). Approaches for identifying ASVs do not cluster sequences based
on a distance-based threshold (11). Proponents of ASVs are largely
dissmissive of concerns that most bacterial genomes have more than one
copy of the \emph{rrn} operon and those copies are not identical (12,
13). Yet, using too fine a threshold to identify OTUs creates the risk
of splitting a single genome into multiple clusters. Conversely, using
too broad of a threshold to define OTUs creates the risk of clustering
together bacterial species into the same OTU. An example of both is seen
in the comparison of \emph{Staphylococcus aureus} (NCTC 8325) and
\emph{S. epidermidis} (ATCC 12228) where each genome has 5 copies of the
16S rRNA gene. Each of the 10 copies of the 16S rRNA gene in these two
genomes is distinct and represent 10 ASVs. Conversely, if the copies
were clustered using a 3\% distance threshold, then all 10 ASVs would
cluster into the same OTU. The goal of this study was to quantify the
risk of splitting a single genome into multiple clusters and the risk of
clustering different bacterial species into the same cluster when using
ASVs and various OTU definitions.

To investigate the variation in the number of copies of the 16S rRNA
gene per genome and the intragenomic variation among copies of the 16S
rRNA gene, I obtained 16S rRNA sequences from the \emph{rrn} copy number
database (\emph{rrn}DB)(14). Among the \textbf{4,774} species
represented in the \emph{rrn}DB there were \textbf{15,614} genomes. The
median \emph{rrn} copy number per species ranged between \textbf{1}
(e.g., \textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
(\textbf{\emph{Metabacillus litoralis}}). As the number of copies of the
operon in a genome increased, the number of variants of the 16S rRNA
gene in each genome also increased. On average, there were \textbf{0.60}
variants per copy of the full length 16S rRNA gene and an average of
\textbf{0.33}, \textbf{0.26}, and \textbf{0.27} variants when
considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
Although a species tended to have a consistent number of 16S rRNA gene
copies per genome, the number of total variants increased with the
number of genomes that were sampled (\textbf{Figure 1}). For example,
the \textbf{180} genome accessions of \textbf{\emph{Mycobacterium
tuberculosis}} in the \emph{rrn}DB each had \textbf{1} copy of the gene
per genome. However, across those accessions, there were \textbf{11}
versions of the gene. An \emph{E. coli} genome typically had \textbf{7}
copies of the 16S rRNA gene with a median of \textbf{5} distinct full
length ASVs per genome (intraquartile range between \textbf{4} and
\textbf{6}). Across the \textbf{958} \emph{E. coli} genomes in the
\emph{rrn}DB, there were \textbf{1,013} versions of the gene. These
observations highlight the risk of selecting a threshold for defining
clusters that is too narrow because it is possible to split a single
genome into multiple clusters.

A method to avoid splitting a single genome into multiple clusters is to
cluster 16S rRNA gene sequences together based on their distances
between each other. Therefore, I assessed the impact of the distance
threshold used to define clusters of 16S rRNA genes on the propensity to
split a genome into separate clusters. To control for uneven
representation of genomes across species, I randomly selected one genome
from each species for the following analyses and repeated each
randomization 100 times. I observed that as the \emph{rrn} copy number
increased, the distance threshold required to reduce the ASVs in each
genome to a single OTU increased (Figure 1). Among species with 7 copies
of the \emph{rrn} operon (e.g., \emph{E. coli}), a distance threshold of
\textbf{5.5}\% was required to reduce full length ASVs into a single OTU
for 95\% of the species. Similarly, thresholds of \textbf{4.0},
\textbf{2.5}, and \textbf{3.5}\% were required for the V3-V4, V4, and
V4-V5 regions, respectively. But, if a 3\% distance threshold was used,
then ASVs from genomes containing fewer than \textbf{5}, \textbf{6},
\textbf{8}, and \textbf{6} copies of the \emph{rrn} operon would
reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4,
V4, and V4-V5 regions, respectively. Consequently, these results
demonstrate that broad thresholds must be used to avoid splitting
different operons from the same genome into separate clusters. At broad
thresholds, 16S rRNA gene sequences from multiple species could be
clustered into the same ASV or OTU (\textbf{Figure 2}). Using ASVs,
\textbf{3.6}\% of the ASVs contained sequences from multiple species
when considering full length sequences and \textbf{10.2}, \textbf{14.9},
and \textbf{12.0}\% when considering the V3-V4, V4, and V4-V5 regions,
respectively. At the commonly used 3\% threshold for defining OTUs,
\textbf{25.2}\% of the OTUs contained 16S rRNA gene sequences from
multiple species when considering full length sequences and
\textbf{29.4}, \textbf{33.0}, and \textbf{32.2}\% when considering the
V3-V4, V4, and V4-V5 regions, respectively. Considering that species
designations are unevenly applied and reflect multiple human-imposed
biases, the risk of splitting a genome into multiple OTUs more
problematic than clustering species together. Therefore, larger
thresholds are advisable.

@@

\textbf{(vi) Note on usage of ASV, OTU, and cluster.} I used ``ASV'' to
denote the cluster of true 16S rRNA gene sequences that were identical
to each other and ``OTU'' to denote the product of distance-based
clustering of sequences. Although ASVs do represent a type of
operational defition of a taxonomic unit and can be thought of as an OTU
formed using a distance of zero, proponents of the ASV approach prefer
to avoid the term OTU given the long history of OTUs being formed by
distance-based clustering
(\url{https://github.com/benjjneb/dada2/issues/62}; accessed
2021-02-19). For this reason, when an ASV splits a genome into different
units those units are called clusters rather than OTUs.

@@

\textbf{Acknowledgements.} I am grateful to Robert Hein and Thomas
Schmidt, who maintain the \emph{rrn}DB, for their help in understanding
the curation of the database and for making the 16S rRNA gene sequences
and related metadata publicly available. I am also grateful to community
members who watched the serialized version of this analysis on YouTube
and provided suggestions and questions over the course of the
development of this project. This work was supported, in part, through
grants from the NIH to PDS (P30DK034933, U01AI124255, and R01CA215574).

@@

\textbf{Figure S1. The ratio of number of distinct ASVs per copy of the
\emph{rrn} operon increased for a species as the number of genomes in
the \emph{rrn}DB for that species increased.} Each point represents a
different species and was shaded to be 80\% transparent so that when
points overlap they become darker. The blue line represents a smoothed
fit through the data. The axes both use a logarithmic scale (base 10).

>>>>

16S rRNA gene sequencing is a powerful technique for describing and
comparing microbial communities (1). Efforts to link 16S rRNA gene
sequences to taxonomic levels based on distance thresholds date to at
least the 1990s. The distance-based threshold that was developed and is
now widely used was based on DNA-DNA hybridization approaches that are
not as precise as genome sequencing (2, 3). Instead, genome sequencing
technologies have suggested that the widely used 3\% distance threshold
to operationally define bacterial taxa is too coarse (4--6). As an
alternative to OTUs, amplicon sequencing variants (ASVs) have been
proposed as a way to adopt the thresholds suggested by genome sequencing
to microbial community analysis using 16S rRNA gene sequences (7--10).
Approaches for identifying ASVs do not cluster sequences based on a
distance-based threshold (11). Proponents of ASVs are largely
dissmissive of concerns that most bacterial genomes have more than one
copy of the \emph{rrn} operon and that those copies are not identical
(12, 13). Yet, using too fine a threshold to identify OTUs creates the
risk of splitting a single genome into multiple clusters. Conversely,
using too broad of a threshold to define OTUs creates the risk of
clustering together bacterial species into the same OTU. An example of
both is seen in the comparison of \emph{Staphylococcus aureus} (NCTC
8325) and \emph{S. epidermidis} (ATCC 12228) where each genome has 5
copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene
in these two genomes is distinct and represent 10 ASVs. Conversely, if
the copies were clustered using a 3\% distance threshold, then all 10
ASVs would cluster into the same OTU. The goal of this study was to
quantify the tradeoff of splitting a single genome into multiple
clusters and the risk of clustering different bacterial species into the
same cluster when using ASVs and various OTU definitions.

To investigate the variation in the number of copies of the 16S rRNA
gene per genome and the intragenomic variation among copies of the 16S
rRNA gene, I obtained 16S rRNA sequences from the \emph{rrn} copy number
database (\emph{rrn}DB)(14). Among the \textbf{4,774} species
represented in the \emph{rrn}DB there were \textbf{15,614} genomes. The
median \emph{rrn} copy number per species ranged between \textbf{1}
(e.g., \textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
(\textbf{\emph{Metabacillus litoralis}}). As \textbf{rrn} copy number
for a genome increased, the number of variants of the 16S rRNA gene in
each genome also increased. On average, there were \textbf{0.60}
variants per copy of the full length 16S rRNA gene and an average of
\textbf{0.33}, \textbf{0.26}, and \textbf{0.27} variants when
considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
Although a species tended to have a consistent number of 16S rRNA gene
copies per genome, the number of total variants increased with the
number of genomes that were sampled (\textbf{Figure 1}). For example,
the \textbf{180} genome accessions of \textbf{\emph{Mycobacterium
tuberculosis}} in the \emph{rrn}DB each had \textbf{1} copy of the gene
per genome. However, across those accessions, there were \textbf{11}
versions of the gene. An \emph{E. coli} genome typically had \textbf{7}
copies of the 16S rRNA gene with a median of \textbf{5} distinct full
length ASVs per genome (intraquartile range between \textbf{4} and
\textbf{6}). Across the \textbf{958} \emph{E. coli} genomes in the
\emph{rrn}DB, there were \textbf{1,013} versions of the gene. These
observations highlight the risk of selecting a threshold for defining
clusters that is too narrow because it is possible to split a single
genome into multiple clusters.

A method to avoid splitting a single genome into multiple clusters is to
cluster 16S rRNA gene sequences together based on their distances
between each other. Therefore, I assessed the impact of the distance
threshold used to define clusters of 16S rRNA genes on the propensity to
split a genome into separate clusters. To control for uneven
representation of genomes across species, I randomly selected one genome
from each species for the following analyses and repeated each
randomization 100 times. I observed that as the \emph{rrn} copy number
increased, the distance threshold required to reduce the ASVs in each
genome to a single OTU increased (Figure 1). Among species with 7 copies
of the \emph{rrn} operon (e.g., \emph{E. coli}), a distance threshold of
\textbf{5.5}\% was required to reduce full length ASVs into a single OTU
for 95\% of the species (Figure 2). Similarly, thresholds of
\textbf{4.0}, \textbf{2.5}, and \textbf{3.5}\% were required for the
V3-V4, V4, and V4-V5 regions, respectively. But, if a 3\% distance
threshold was used, then ASVs from genomes containing fewer than
\textbf{5}, \textbf{6}, \textbf{8}, and \textbf{6} copies of the
\emph{rrn} operon would reliably be clustered into a single OTU for ASVs
from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 2).
Consequently, these results demonstrate that broad thresholds must be
used to avoid splitting different operons from the same genome into
separate clusters. At broad thresholds, 16S rRNA gene sequences from
multiple species could be clustered into the same ASV or OTU
(\textbf{Figure 2}). Using ASVs, \textbf{3.6}\% of the ASVs contained
sequences from multiple species when considering full length sequences
and \textbf{10.2}, \textbf{14.9}, and \textbf{12.0}\% when considering
the V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3\%
threshold for defining OTUs, \textbf{25.2}\% of the OTUs contained 16S
rRNA gene sequences from multiple species when considering full length
sequences and \textbf{29.4}, \textbf{33.0}, and \textbf{32.2}\% when
considering the V3-V4, V4, and V4-V5 regions, respectively. Considering
that species designations are unevenly applied and reflect multiple
human-imposed biases, the risk of splitting a genome into multiple OTUs
is more problematic than clustering species together. Therefore, larger
thresholds are advisable.

@@

\textbf{(vi) Note on usage of ASV, OTU, and cluster.} I used ``ASV'' to
denote the cluster of true 16S rRNA gene sequences that were identical
to each other and ``OTU'' to denote the product of distance-based
clustering of sequences. Although ASVs do represent a type of
operational defition of a taxonomic unit and can be thought of as an OTU
formed using a distance of zero, proponents of the ASV approach prefer
to avoid the term OTU given the long history of OTUs being formed by
distance-based clustering
(\url{https://github.com/benjjneb/dada2/issues/62}; accessed
2021-02-19). For this reason, when an ASV splits a genome into different
units, those units were called clusters rather than OTUs.

@@

\textbf{Acknowledgements.} I am grateful to Robert Hein and Thomas
Schmidt, who maintain the \emph{rrn}DB, for their help in understanding
the curation of the database and for making the 16S rRNA gene sequences
and related metadata publicly available. I am also grateful to community
members who watched the serialized version of this analysis on YouTube
and provided suggestions and questions over the course of the
development of this project. This work was supported, in part, through
grants from the NIH (P30DK034933, U01AI124255, and R01CA215574).

@@

\textbf{Figure S1. The ratio of number of distinct ASVs per copy of the
\emph{rrn} operon increased for a species as the number of genomes in
the \emph{rrn}DB for that species increased.} Each point represents a
different species and was shaded to be 80\% transparent so that when
points overlap they become darker. The blue line represents a smoothed
fit through the data. Both axes use a logarithmic scale (base 10).
